Fast time-domain electromagnetic modelling in the frequency domain

### 2020-06-22 ManuscriptTime.tex
Improve environment and instructions

% Own commands
\newcommand{\mr}[1]{\mathrm{#1}}
\newcommand{\emg}[2]{\texttt{emg#1#2}\xspace}
\newcommand{\empymod}{\texttt{empymod}\xspace}
\newcommand{\simpeg}{\texttt{SimPEG}\xspace}

@@

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft;\\
\footnotemark[2]Shell Global Solutions International BV,
Grasweg 31, 1031 HW Amsterdam;\\
E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org};\\
\textbf{Keywords}: CSEM, Fourier transforms, 3D modelling.
}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Modelling three-dimensional time-domain controlled-source electromagnetic data
with frequency-domain solvers is an alternative to time-domain solvers. It
requires the computation of many frequencies if standard Fourier transforms are
used, which can make it prohibitively expensive in comparison with a
time-domain computation. The speed of time-domain modelling with
frequency-domain computation is defined by three key points: solver, method and
implementation of the Fourier transformation, and gridding.
%
The faster the frequency-domain solver, the faster time-domain modelling will
be. It is important that the solver is robust over a wide range of frequencies.
A three-dimensional, iterative, frequency-domain multigrid solver satisfies
this requirement, but any other robust solver will do.
%
The method should require as few frequencies as possible while remaining
robust. As the frequency range spans many orders of magnitude, the required
frequencies are ideally equally spaced on a logarithmic scale, as is the case
for, e.g., the digital linear filter method or the logarithmic fast Fourier
transform. With careful selection of frequencies and interpolation, 15 to 20
frequencies are enough to cover a wide range of offsets in one-dimensional test
models.
%
The gridding should be frequency-dependent, which is generally accomplished by
making it a function of skin depth. Optimising for the least number of required
cells for a given frequency should be combined with optimising for
computational speed.
%
Looking carefully at these three points results in much faster computation
times with speedup-factors of ten and more over previous results. A
frequency-domain computation of time-domain data can therefore be competitive
to time-domain solvers if the required frequencies and their corresponding
grids are carefully chosen.
%
\end{abstract}

@@

Acquired CSEM data are often analysed (modelled and inverted for) in their
respective domain, either the frequency or time domain. Modellers of layered
media usually exploit the horizontal shift-invariance by computing the
responses in the wavenumber-frequency domain followed by a 2D inverse spatial
Fourier transform, also called Hankel transform, to the space-frequency domain,
and a regular inverse Fourier transform if time-domain data are required
\citep[e.g., ][]{GEO.15.Hunziker}.

CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time. However,
under certain conditions it can be competitive to model time-domain data with a
frequency-domain code, as shown by \cite{GEO.08.Mulder}. The conditions are: a
sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are an adaptive, frequency-dependent gridding
scheme that minimizes the required cells in each dimension, and a logarithmic
Fourier transform such as digital linear filters \citep[DLF, ][]{GP.71.Ghosh}
or the logarithmic fast Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton}
to go from the frequency to the time domain. The latter makes it also possible
to only use the imaginary part of the frequency-domain response, which has
advantages when it comes to interpolation.

@@

Time-domain modelling with a frequency-domain code requires the computation of
a relatively wide range of frequencies for the Fourier transform. However, the
required computation grids for low frequencies (in our case in the order of
0.001\,Hz) and high frequencies (around 100\,Hz) are hugely different. Low
frequencies can be computed on a coarser grid, but they require a much larger
domain in order to avoid boundary effects. High frequencies, on the other hand,
require denser gridding, but they are much more limited in reach. An adaptive
gridding scheme is therefore required. We build our approach upon
\cite{GEO.07.Plessix}, who presented such an adaptive gridding for
multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain is naturally based on the skin depth,
which is the distance after with the amplitude of the electromagnetic field is
decayed by $1/e\approx 37\,\%$. The skin depth $\delta$ is a function of
conductivity and frequency, and for the diffusive approximation of Maxwell's
equation in an isotropic, homogeneous medium is given by \citep[e.g.,][
equation~1.53]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \
  % \frac{503.3}{\sqrt{f\sigma}} \, ,
  % 503.3(f\sigma)^{-1/2} \, ,
  503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

@@

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code even more competitive in comparison with time-domain
codes.

@@

For the comparison we use the open-source code \simpeg \citep{CAG.15.Cockett}.
\simpeg is a framework for modelling and inversion of geophysical data such as
gravity, magnetics, and CSEM. It has Maxwell's equations implemented in both
the frequency and time domain. As such we can compare our result computed in
the frequency domain followed by a Fourier transform to a result computed
directly in the time domain. A principal difference between \simpeg and \emg3d
is that the former has various direct solvers implemented, whereas the latter
is an iterative multigrid solver. The 3D model is therefore a rather small
example in order to be able to run it on our test machine, as the memory
requirement by the direct solver would otherwise be too high.

@@

We have shown that time-domain CSEM modelling with a frequency-domain code can
be competitive given a robust frequency-domain solver, a frequency-dependent
gridding function that minimizes the required cells, and a Fourier transform
that works on a logarithmic scale. Fast layered modelling can be used to design
the required frequency range, as the Fourier transform does not know about the
dimensionality of the underlying model. Twenty frequencies or less are usually
sufficient for a wide range of offsets. We believe that our proposed
improvements to the previously published methods make time-domain CSEM
modelling with a frequency-domain code even more attractive than it was already
before.

>>>>

% Own commands
\newcommand{\mr}[1]{\mathrm{#1}}
\newcommand{\emg}[2]{\texttt{emg#1#2}\xspace}
\newcommand{\empymod}{\texttt{empymod}\xspace}
\newcommand{\simpeg}{\texttt{SimPEG}\xspace}
\newcommand{\custem}{\texttt{custEM}\xspace}

@@

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft;\\
\footnotemark[2]Shell Global Solutions International BV,
Grasweg 31, 1031 HW Amsterdam;\\
E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org};\\
\textbf{Keywords}: Fourier transforms, CSEM, 3D modelling.
}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Modelling three-dimensional time-domain controlled-source electromagnetic data
with frequency-domain solvers is an alternative to time-domain solvers. It
requires the computation of many frequencies if standard Fourier transforms are
used, which can make it prohibitively expensive in comparison with a
time-domain computation. The speed of time-domain modelling with
frequency-domain computation is defined by three key points: solver, method and
implementation of the Fourier transformation, and gridding.
%
The faster the frequency-domain solver, the faster time-domain modelling will
be. It is important that the solver is robust over a wide range of frequencies.
A three-dimensional, iterative, frequency-domain multigrid solver satisfies
this requirement, but any other robust solver will do.
%
The method should require as few frequencies as possible while remaining
robust. As the frequency range spans many orders of magnitude, the required
frequencies are ideally equally spaced on a logarithmic scale. We present a
fast frequency-to-time domain method using either the digital linear filter
method or the logarithmic fast Fourier transform together with careful
selection of frequencies and interpolation. 15 to 20 frequencies are generally
enough with this methodology to cover a wide range of offsets in
one-dimensional test models.
%
The gridding should be frequency-dependent, which is accomplished by making it
a function of skin depth. Optimising for the least number of required cells for
a given frequency should be combined with optimising for computational speed.
%
Looking carefully at these three points results in much faster computation
times with speedup-factors of ten and more over previous results. A
frequency-domain computation of time-domain data can therefore be competitive
to time-domain solvers if the required frequencies and their corresponding
grids are carefully chosen.
%
\end{abstract}

@@

CSEM is usually divided into frequency- and time-domain methods, depending
whether the source signal is a continuous waveform, such as a sine, or a finite
waveform, such as a pseudo-random binary sequence (PRBS). A numerical
comparison of the two methods is given in \cite{GP.13.Conell}. Acquired CSEM
data are subsequently often analysed (modelled and inverted for) in their
respective domain, either the frequency or time domain. Modellers of layered
media usually exploit the horizontal shift-invariance by computing the
responses in the wavenumber-frequency domain followed by a 2D inverse spatial
Fourier transform, also called Hankel transform, to the space-frequency domain,
and a regular inverse Fourier transform if time-domain data are required
\citep[e.g., ][]{GEO.15.Hunziker}.

CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time (see,
e.g., \cite{SG.10.Borner, SG.05.Avdeev} for comprehensive overviews of CSEM
codes). However, under certain conditions it can be competitive to model
time-domain data with a frequency-domain code, as shown by
\cite{GEO.08.Mulder}. The conditions are: a sufficiently powerful solver,
appropriate frequency selection and interpolation, and an automated gridding,
for which they used the multi-frequency CSEM approach presented by
\cite{GEO.07.Plessix}. We build upon these results but improve the run time
from hours to minutes. The main reasons for this significant speed-up are an
adaptive, frequency-dependent gridding scheme that minimizes the required cells
in each dimension, and a logarithmic Fourier transform such as digital linear
filters \citep[DLF, ][]{GP.71.Ghosh} or the logarithmic fast Fourier transform
\citep[FFTLog, ][]{RAS.00.Hamilton} to go from the frequency to the time
domain. The latter makes it also possible to only use the imaginary part of the
frequency-domain response, which has advantages when it comes to interpolation.

@@

Being able to model CSEM data both in the frequency-domain and in the
time-domain can be desirable, as both domains have advantages and
disadvantages. One way to achieve this is to implement Maxwell's equation in
both domains, as it is done, for instance, in \simpeg \citep{CAG.15.Cockett}.
Another method is to have Maxwell's equation only implemented in one domain,
and use Fourier transforms to go to the other domain. However, this approach
can be costly, as many frequencies over a wide range are required to go from
the time domain to the frequency domain, or many times over a wide range for
the opposite direction. We present a methodology which significantly reduces
the amount and range of the required frequencies, which translates into a
significant faster computation times. The required computation grids for low
frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain is naturally based on the skin depth,
which is the distance after which the amplitude of the electromagnetic field is
decayed by $1/e\approx 37\,\%$. The skin depth $\delta$ is a function of
conductivity and frequency, and for the diffusive approximation of Maxwell's
equation in an isotropic, homogeneous medium is given by \citep[e.g.,][
equation~1.53]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \
  % \frac{503.3}{\sqrt{f\sigma}} \, ,
  % 503.3(f\sigma)^{-1/2} \, ,
  503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

@@

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code even more competitive in comparison with time-domain
codes. It is important to note that while we show the transform from frequency
to time domain, the conclusions regarding the Fourier transform can equally
well be applied to the transform from time to frequency domain data for a code
that solves Maxwell's equations in the time domain. A variation of our method
was implemented, for instance, by \custem \cite{GEO.19.Rochlitz} to obtain
frequency-domain data from time-domain computations (Rochlitz, personal
communication).

@@

For the comparison we use the open-source code \simpeg \citep{CAG.15.Cockett}.
\simpeg is a framework for modelling and inversion of geophysical data such as
gravity, magnetics, and CSEM. It has Maxwell's equations implemented in both
the frequency and time domain. As such we can compare our result computed in
the frequency domain followed by a Fourier transform to a result computed
directly in the time domain. A principal difference between \simpeg and \emg3d
is that the former has various direct solvers implemented, whereas the latter
is an iterative multigrid solver. The 3D model is therefore a rather small
example in order to be able to run it on our test machine, as the memory
requirement by the direct solver would otherwise be too high. There are not
many options out there of open-source time-domain 3D codes, \simpeg being the
one we found to be suitable. A step-off response was chosen as this is the
response currently implemented in \simpeg.

@@

We have shown a method to minimize the required frequencies and their range for
the computation of time-domain CSEM data with a frequency-domain code. This can
significantly reduce the computation time and makes time-domain CSEM modelling
with a frequency-domain code competitive given a robust frequency-domain
solver, a frequency-dependent gridding function that minimizes the required
cells, and a Fourier transform that works on a logarithmic scale. Fast layered
modelling can be used to design the required frequency range, as the Fourier
transform does not know about the dimensionality of the underlying model.
Twenty frequencies or less are usually sufficient for a wide range of offsets.
We believe that our proposed improvements to the previously published methods
make time-domain CSEM modelling with a frequency-domain code even more
attractive than it was already before. We have demonstrated the idea of our
Fourier transform method on CSEM data transformed from the frequency to the
time domain. However, it could equally be applied to the transform from the
time to the frequency domain and to other methods with similar
characteristics.
